The brightness and spatial distributions of terrestrial radio sources

### 2013-07-16 articles/Article-RFI-distribution/article-rfi-distribution.tex
Removed all editmarks and extra comments

\usepackage[colorlinks=true,linkcolor=customlinkcolor,urlcolor=customlinkcolor,citecolor=customcitecolor,pdftex]{hyperref}

@@

%\newcommand{\editmark}[1]{#1}
\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}

@@

In this article, we will connect new insights about RFI to the angular \editmark{EoR} experiment that is using the Low-Frequency Array (LOFAR) (\citealt{de-bruyn-eor-ursi-2011}, \citealt{lofar-overview-2013}). The LOFAR EoR project aims to detect the redshifted 21-cm signals from the EoR using the LOFAR HBA antennas (115--190~MHz, $z_\textrm{HI}$=11.4--6.5). Several fields will be observed over 100~nights, to achieve sufficient sensitivity to allow the signal extraction. An EoR calibration pipeline has been designed that solves for ionospheric and instrumental effects in approximately hundred directions using the SAGE algorithm \citep{sage-calibration-ii}. Initial results from commissioning observations show that in a single night the thermal noise level can almost be reached \citep{ncp-eor-yatawatta}.
% TODO correct reference if van Haarlem et al is published

@@

This paper is organised as follows: in Sect.~\ref{sec:distribution-prediction}, we calculate a model for terrestrial interfering source distribution based on various assumptions. Sect.~\ref{sec:distribution-methods} presents the methods that we use to generate and analyse brightness histograms of LOFAR data. \editmark{Sect.~\ref{sec:dist-data} describes} the two LOFAR data sets that have been used to perform the experiment. The results of analysing the sets are presented in Sect.~\ref{sec:dist-results}. Finally, in Sect.~\ref{sec:dist-discussion} the results are discussed and conclusions are drawn.

@@

It is common knowledge that in a uniform Euclidian Universe source counts behave like power-law distributions. \editmark{The differential distribution for sources on a flat surface is a power law with -2 exponent}. We will derive this expected intrinsic source distribution for interfering radio sources. After that, we will analyse the issues that arise when measuring the distribution by counting samples.

@@

Next, we assume that all RFI sources have equal constant strength $I$ and follow a uniform spatial distribution in the local two-dimensional horizontal plane. \editmark{These assumptions are obviously simplications, but we will address these later. %This is obviously a simplification, as the RFI sources are actually distributed on the surface of a sphere with different heights. Beyond some distance, the Earth will partly block the path between transmitter and \editmark{receiver.
Using} these assumptions, we can express the expected \editmark{inverse cumulative rate density of sources at distance $r$ as
\begin{equation} \label{eq:inverse-cumulative-distance-function}
%F_{\textrm{distance} \le r}(r)=\rho \pi r^2, (<-- non inverted)
F_{\textrm{distance} \ge r}(r) = N - \rho \pi r^2,
\end{equation}
with $N$ the total number of sources and for some constant $\rho$ that represents the number of sources per unit area.}
%In other words, we will observe $F$ sources that are further than a distance $r$ away. We need $F_{\textrm{distance} \ge r}(r)$, the complement of Eq.~\eqref{eq:cumulative-distance-function} to calculate the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance. Sources with at least a distance of $r$ are given by $F_{\textrm{distance} \ge r}(r) = N - \rho \pi r^2$, with $N$ the total number of sources. Because $r$ is restricted, $N$ is finite.
The cumulative number of sources $F_{\textrm{amplitude} \le S}$ that have an amplitude of at most $S$ can \editmark{be calculated from this with}
\begin{align} \label{eq:cumulative-rfi-distribution}
 F_{\textrm{amplitude} \le S}(S) = & F_{\textrm{distance} \ge r}(\mathcal{R}(S))
%\notag = & F_{\textrm{distance} \ge r}(\pm\sqrt\frac{Ig}{4\pi S})\\
 = N-\frac{\rho Ig}{4S}
\end{align}
where $\mathcal{R}(S)=\mathcal{S}^{-1}$, the inverse of $\mathcal{S}$, i.e., the function that returns the distance $r$ for a given amplitude $S$. Finally, the rate density can be calculated by taking the derivative,
\begin{align} \label{eq:two-dimensional-distribution}
 f_S(S) = & \frac{dF_{\textrm{amplitude} \le S}}{dS} = \frac{\rho Ig}{4S^2}.
\end{align}

Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we expect to see a power law with a slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional \editmark{plane.} %For this, we have assumed that the variables $I$, $g$ and $\rho$ are constants. In reality, these variables have a stochastic nature. The effect of this will be discussed in \S\ref{sec:parameter-variability}.

%\subsection{Spherical case} \label{sec:spherical-geometry}
%If we do take into account the fact that the sources lie on a sphere instead of a plane, Eq.~\eqref{eq:cumulative-distance-function} changes. If we assume their difference in height is negligible, we can use the formula for the surface area of a spherical cap,
%\begin{equation}\label{eq:spherical-cap-area}
% A_\textrm{cap} = \pi \left( a^2 + h^2 \right) \hspace{0.4cm}\textrm{with } (a \le R) \wedge (h \le 2R),
%\end{equation}
%with $R$ the radius of the Earth, $a$ the radius of the base of the cap and $h$ the height of the cap. Because the Euclidean distance $r$ between the source and the antenna is $r^2 = a^2 + h^2$, the resulting number of sources within Euclidean distance $r$ becomes
%\begin{equation} \label{eq:cumulative-spherical}
%\tilde{F}_{\textrm{distance} \le r}(r) = \rho A_\textrm{cap} = \rho \pi r^2 \hspace{0.4cm}\textrm{with } r \le 2R,
%\end{equation}
%which equals the formula for a planar source distribution given in Eq.~\eqref{eq:cumulative-distance-function}. Therefore, the spherical and two-dimensional cases lead to a distribution with equal power law. Eq.~\eqref{eq:cumulative-spherical} is valid for $r \le 2R$, but in practice the lowest amplitude of observed sources is further constrained because of the curvature of the Earth. This causes distant sources to settle below the horizon. Nevertheless, some of those sources might still be visible due to reflection of the ionosphere and other (semi-)spherical propagation effects.

\subsection{Propagation effects}
So far, we have assumed that the electromagnetic radiation propagates through free space, resulting in an $r^{-2}$ fall-off. In reality, the radiation will be affected by complicated propagational effects \editmark{due to the surface of the Earth}. %Because of the irregular surface of the Earth (including urban areas) and the absence of a direct line of sight between transmitter and receiver, the propagation mode might be indirect. Multiple indirect paths might be formed that include reflection, refraction, diffraction and absorption of the electromagnetic wave. 
A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}. %The original model is based on urban areas, but corrections are given by Hata for suburban areas with a lower population density and for open areas.
\editmark{Hata} gives the following analytical estimate for $L_p$, the electromagnetic propagation loss between two ground-based antennas:
\begin{align} 
\notag                L_p & = & 69.55 + 26.16 \log_{10} f_c - 13.82 \log_{10} h_b - \\
\label{eq:hata-model} & &  a(h_m) + (44.9 - 6.55 \log_{10} h_b) \log_{10} r, 
\end{align}
where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in meters; $h_m$ the height of the receiving antenna in meters; $r$ the distance between the antennas in meters; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the urban density. \citeauthor{hata-propagation-loss} found this model to be representative for frequencies $f_c \sim$~150--1500~MHz, with transmitter heights $h_b \sim$~30--200~m, receiver heights $h_m \sim$~1--10~m and over distances $r\sim$~1--20~km.

Converting from a subtracted term in decibels to a flux density factor $L_S$ \editmark{results in}
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \frac{1}{10} 10^{L_p} = \zeta r^\eta,
\end{equation}
\editmark{with $\eta$ and $\zeta$ given by}
\begin{align} 
 \label{eq:zeta-definition} \eta & = 4.49 - 0.655 \log_{10} h_b,\\
 \label{eq:eta-definition}  \zeta & = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{align}
Note that according to Hata's model, the exponent of the power law $\eta$ depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. To find the rate density function $f_p$ that considers propagation effects, one can replace $S(r)$ in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{Ig}{4\pi \zeta r^\eta}.
\end{equation}
\editmark{The} resulting rate density function $f_p$ is
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(rho*%pi*r^2,S),solve(S(r)=S,r);
\begin{align} \label{eq:frequency-density-with-propagation}
f_p(S)
= & \frac{d}{dS} \left[ N - \rho \pi \left( \frac{Ig }{4\pi\zeta S} \right)^{2/\eta} \right]
= \frac{\rho 2\pi}{\eta S} \left( \frac{Ig}{4\pi\zeta S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free-space propagation effects, the observed log-log histogram is predicted to have a $-(\frac{2}{\eta}+1)$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram. According to Hata's model, this is valid for the range 30--200 m. The trend of the slope will not continue indefinitely when increasing the height further. Instead it will converge to a $-2$ slope, which corresponds to free-space propagation.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
\editmark{This} yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

\editmark{We note that a uniform distribution of meteors or aircrafts which reflect free-space propagating RFI can create a power-law distribution with a similar slope: a uniform two-dimensional distribution of reflecting sources will create a $-1.5$ slope, while a uniform three-dimensional distribution will create a $-1.75$ slope. With brightness-distribution analyses one can therefore not distinguish between transmitters affected by Hata's propagation model and reflectors affected by free-space propagation. Reflected RFI might become relevant at lower amplitude levels.}

@@

So far, these are the expected histograms for pure noise and pure RFI that propagates through free space. However, the measured distribution is a mixture of the \editmark{two. %A perfect RFI detector would classify the samples in two groups; one group of samples that are not affected by RFI, and therefore contain noise only, and one group of samples that are contaminated, and are the sum of RFI and noise. Real RFI detectors can classify samples into these groups to some extent, but due to false positives and false negatives, the groups will get mixed nevertheless. Even more, we still need to take into account that the samples classified as RFI are also affected by noise. Although this effect is moderate, because `most' RFI samples are of much higher amplitude than what is added because of the noise, we are interested in the low-level RFI samples, and the relative effect of noise on those samples is large.
Analytic derivation of the corresponding mixed amplitude distribution is not trivial, but the distributions can easily be} estimated by drawing complex samples from the two distributions and calculating and counting the amplitudes. A sample can be drawn from the RFI distribution by integration, scaling and inversion of the rate density function in Eq.~\eqref{eq:frequency-density-with-propagation}. %Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of RFI sources will go to infinity.
\editmark{To invert the cumulative function, one needs to assume that there are no sources beyond some limiting distance $r_L$}. With this assumption, a single complex RFI contaminated sample $S_{\textrm{RFI}}$ can be sampled with:
\begin{equation} \label{eq:drawing-rfi-with-propagation}
% === Proof of formula below ===
% From definition above:
%   F(S) = N-\rho \pi ( Ig / 4 \pi \zeta S ) ^ (2/\eta)
% We want S to be in the interval [ S_l ; \inf ], where: (from equation of S(r) above)
%   S_l = I g / 4 \pi \zeta r_l^\eta,
% the flux density of the most distance sources at distance r_l that are still visible.
% This means F(S) will be in the interval [ F(\inf) ; F(S_l) ], and:
%   F(\inf) = N
%    F(S_l) = N - \rho \pi r_l^2
% Because F is the CDF, we can sample a strength by calculating the S
% from a uniformly distributed F within this interval.
% If x_u is a uniform variable between 0 and 1, then N - x_u \rho \pi r_l^2 is a uniform
% variable in F's interval. From this we calculate S with the definition of F(S) :
%   N - x_u \rho \pi r_l^2 = N-\rho \pi ( Ig / 4 \pi \zeta S ) ^ (2/\eta)
%       x_u \rho \pi r_l^2 = \rho \pi ( Ig / 4 \pi \zeta S ) ^ (2/\eta)
%                x_u r_l^2 = ( Ig / 4 \pi \zeta S ) ^ (2/eta)
%   (x_u r_l^2 ) ^ (eta/2) = Ig / 4 \pi \zeta S
 %   x_u^(\eta/2) r_l^\eta = Ig / 4 \pi \zeta S
%                        S = Ig / ( 4 \pi \zeta x_u^(\eta/2) r_l^\eta )
S_{\textrm{RFI}} \leftarrow \frac{I g}{4\pi \zeta x_u^{\eta/2} r_L^\eta} e^{i 2 \pi y_u}.
\end{equation}
Here, $S_{\textrm{RFI}}$ is a new \editmark{complex RFI sample that follows a power-law distribution; $\eta$ and $\zeta$ are defined in Eqs.~\eqref{eq:zeta-definition} and \eqref{eq:eta-definition};} $I$ is the average intrinsic strength; $g$ is the gain of the instrument; $0 < x_u,y_u \le 1$ are two independently drawn uniformly distributed samples; and $r_L$ is the maximum distance of visible sources.
% Free-space propagating drawing formula:
%\begin{equation} \label{eq:drawing-rfi-distribution}
%S_{\textrm{RFI}} \leftarrow \frac{Ig}{4\pi x_u r_L^2} e^{i 2 \pi y_u},
%\end{equation}
%where $0 < x_u, y_u \le 1$ are independently drawn from a uniform distribution. The result is an amplitude sample $S_{\textrm{RFI}}$, with $S_{\textrm{RFI}} \ge Ig/4\pi r_L^2$, that follows a power-law distribution.
A sample $S$ that is contaminated by both RFI and noise \editmark{can be drawn with
$S \leftarrow v_n + w_n i + S_{\textrm{RFI}}$, with $v_n, w_n \sim N(\mu=0; \sigma)$. An} example of distribution curves of $S$ for $\eta=2$ and various settings of $Ig/4\pi \zeta r_L^2$ is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.

%If we assume that $I$, the average RFI source strength; $g$, the average instrumental gain; and $r_L$, the distance over which those sources are visible, are all unknown, it can be seen from Eq.~\eqref{eq:drawing-rfi-with-propagation} that given the histogram we can solve neither $r_L$, $I$ or $g$, as they all have the same effect of scaling the -2 power-law part of the histogram. Although calibration could in theory solve $g$, almost all RFI sources will come in through the far-sidelobes of the beam, and finding the expected values for the gains is therefore hard. The effects of these parameters on the histogram will be further discussed in \S\ref{sec:parameter-variability}.

%Fig.~\ref{fig:rayleigh-and-rfi-combined} shows that the left side of the graph corresponds with the original Rayleigh distribution. From the right side of the histogram, we are able to estimate $Ig/r_L^2$, as the placing of the power-law curve is independent of the Rayleigh distribution for large $S$. Methods to constrain the parameters of the RFI distribution will be discussed in \S\ref{sec:rfi-distribution-constraints}. The curves in Fig.~\ref{fig:rayleigh-and-rfi-combined} are from histograms of samples that are all contaminated by both noise and RFI. In practice, we can not make this distinction accurately, as RFI detectors have a limited accuracy. Consequently, the observed histogram will be the sum of two types of histograms: the first histogram being contaminated by both noise and RFI, the second one only by noise.

@@

\editmark{From Eq.~\eqref{eq:frequency-density-with-propagation} it can be seen that the $\rho$, $I$ and $g$ parameters have the same effect of scaling the power-law distribution, and do not change its shape or slope. Therefore, with distribution analyses one can e.g. not determine whether the distribution is dominated by low-power sources within the horizon or by scattered signals from over the horizon. The horizon of an antenna is estimated 	with $\sqrt{2hr}$ \citep{bullington-propagation}, with $r$ the radius of the Earth and $h$ the height of the antenna. For LOFAR, the horizon is at about 5~km.}

@@

Fitting straight lines to the distribution curve in a log-log plot is not the most accurate way of estimating the exponent of a power-law distribution \citep{power-law-distribution}. However, because of our enormous sample size, which allows fitting the line over a large interval, the estimator will be sufficiently accurate for our purpose. Nevertheless, we will additionally calculate a maximum-likelihood estimator for comparison. The maximum-likelihood estimator for the exponent in a power-law distribution is given by the Hill estimator $\hat \alpha_H$ \citep{hill-estimator, power-law-distribution}, defined as:
\begin{equation} \label{eq:hill}
 \hat \alpha_H = 1 + N \left(\sum\limits_{i=1}^{N} \ln \frac{x_i}{x_\textrm{min}} \right)^{-1},
\end{equation}
with $N$ the number of samples and $x_i$ for $0 < i \le N$ the samples that follow a power law with \editmark{lower bound $x_\textrm{min}$.}
%However, this estimator assumes the distribution is not cut-off at a high point. In our case, the distribution is cut-off at both sides, for example because of the limit of the analogue-digital converter (see \S\ref{sec:rfi-distribution-constraints}). Therefore, using this estimator will result in an estimate that is steeper (i.e., more negative) than the actual distribution.

\begin{figure}
\begin{center}
%\includegraphics[width=7.5cm]{img/explanation-lower-constraint/FindingCounts}
\includegraphics[width=7.5cm]{img/explanation-lower-constraint/FindingConstraints}
\caption{Cartoon of how a constraint on the lower fall-off point of the power-law distribution can be determined. Note that the labelled areas are areas as occupied in a linear plot, i.e., the integration of the density function. Areas in a log-log plot are not linearly related to the integral. There are two ways to estimate the lower constraint $S_L$: (i) the areas $N_a$ and $N_\textrm{total}-N_\textrm{part}$ are equal if $Ig/r_L^\eta$ is constant during the observation, and (ii) if one assumes $Ig/r_L^\eta\sim$ uniform, then $N_a+N_b = N_\textrm{total}-N_\textrm{part}$.}
\label{fig:explanation-constraints}
\end{center}
\end{figure}

\subsection{Determining RFI distribution limits} \label{sec:rfi-distribution-constraints}
\editmark{In this section we will show how to put upper and lower constraints on the power-law distribution.}
%show methods to find $S_U$ and $S_L$, the upper and lower flux limits of the power-law distribution at which the power law breaks down. Once the exponent of the power-law part of the distribution is estimated with the previously discussed techniques, the distribution can be extrapolated to find a lower flux limit.
Assume that we have found a power law with exponent $\alpha$ and factor $\beta$ over an amplitude region $[S_1 ; S_2]$, resulting in the \editmark{rate density function $h(S) = \beta S^\alpha$}. $S_1$ and $S_2$ are selected by visual inspection of the histogram. Assume that the histogram contains $N_\textrm{part}$ (RFI) samples with amplitude $> S_1$, as sketched in Fig.~\ref{fig:explanation-constraints}, and that the effect of the Rayleigh component on the histogram $> S_1$ is negligible. The hypothetical upper limit $S_U$ of the distribution can be \editmark{found by} solving
\begin{equation} \label{eq:upper-limit-constrained}
 \int\limits_{S_1}^{S_U} h(S) dS = N_\textrm{part}.
\end{equation}
The observed histogram will break down beyond some amplitude because \editmark{of several reasons: the samples are digitized with an analogue-to-digital converter (ADC) with limited range; we observe for a limited time and the rate count is discrete; and, under the assumption of a uniform spatial distribution of RFI transmitters, samples with} very high amplitude would have to be produced by transmitters that are very close to the telescope. However, it is likely that the uniform spatial distribution of transmitters will break down at closer distances.

>>>>

%\usepackage[colorlinks=true,linkcolor=customlinkcolor,urlcolor=customlinkcolor,citecolor=customcitecolor,pdftex]{hyperref}
\usepackage[pdftex]{hyperref}

@@

%\newcommand{\editmark}[1]{#1}
%\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}

@@

In this article, we will connect new insights about RFI to the angular EoR experiment that is using the Low-Frequency Array (LOFAR) (\citealt{de-bruyn-eor-ursi-2011}, \citealt{lofar-overview-2013}). The LOFAR EoR project aims to detect the redshifted 21-cm signals from the EoR using the LOFAR HBA antennas (115--190~MHz, $z_\textrm{HI}$=11.4--6.5). Several fields will be observed over 100~nights, to achieve sufficient sensitivity to allow the signal extraction. An EoR calibration pipeline has been designed that solves for ionospheric and instrumental effects in approximately hundred directions using the SAGE algorithm \citep{sage-calibration-ii}. Initial results from commissioning observations show that in a single night the thermal noise level can almost be reached \citep{ncp-eor-yatawatta}.
% TODO correct reference if van Haarlem et al is published

@@

This paper is organised as follows: in Sect.~\ref{sec:distribution-prediction}, we calculate a model for terrestrial interfering source distribution based on various assumptions. Sect.~\ref{sec:distribution-methods} presents the methods that we use to generate and analyse brightness histograms of LOFAR data. Sect.~\ref{sec:dist-data} describes the two LOFAR data sets that have been used to perform the experiment. The results of analysing the sets are presented in Sect.~\ref{sec:dist-results}. Finally, in Sect.~\ref{sec:dist-discussion} the results are discussed and conclusions are drawn.

@@

It is common knowledge that in a uniform Euclidian Universe source counts behave like power-law distributions. The differential distribution for sources on a flat surface is a power law with -2 exponent. We will derive this expected intrinsic source distribution for interfering radio sources. After that, we will analyse the issues that arise when measuring the distribution by counting samples.

@@

Next, we assume that all RFI sources have equal constant strength $I$ and follow a uniform spatial distribution in the local two-dimensional horizontal plane. These assumptions are obviously simplications, but we will address these later. Using these assumptions, we can express the expected inverse cumulative rate density of sources at distance $r$ as
\begin{equation} \label{eq:inverse-cumulative-distance-function}
%F_{\textrm{distance} \le r}(r)=\rho \pi r^2, (<-- non inverted)
F_{\textrm{distance} \ge r}(r) = N - \rho \pi r^2,
\end{equation}
with $N$ the total number of sources and for some constant $\rho$ that represents the number of sources per unit area. The cumulative number of sources $F_{\textrm{amplitude} \le S}$ that have an amplitude of at most $S$ can be calculated from this with
\begin{align} \label{eq:cumulative-rfi-distribution}
 F_{\textrm{amplitude} \le S}(S) = & F_{\textrm{distance} \ge r}(\mathcal{R}(S))
%\notag = & F_{\textrm{distance} \ge r}(\pm\sqrt\frac{Ig}{4\pi S})\\
 = N-\frac{\rho Ig}{4S}
\end{align}
where $\mathcal{R}(S)=\mathcal{S}^{-1}$, the inverse of $\mathcal{S}$, i.e., the function that returns the distance $r$ for a given amplitude $S$. Finally, the rate density can be calculated by taking the derivative,
\begin{align} \label{eq:two-dimensional-distribution}
 f_S(S) = & \frac{dF_{\textrm{amplitude} \le S}}{dS} = \frac{\rho Ig}{4S^2}.
\end{align}

Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we expect to see a power law with a slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional plane.

\subsection{Propagation effects}
So far, we have assumed that the electromagnetic radiation propagates through free space, resulting in an $r^{-2}$ fall-off. In reality, the radiation will be affected by complicated propagational effects due to the surface of the Earth. A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}. Hata gives the following analytical estimate for $L_p$, the electromagnetic propagation loss between two ground-based antennas:
\begin{align} 
\notag                L_p & = & 69.55 + 26.16 \log_{10} f_c - 13.82 \log_{10} h_b - \\
\label{eq:hata-model} & &  a(h_m) + (44.9 - 6.55 \log_{10} h_b) \log_{10} r, 
\end{align}
where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in meters; $h_m$ the height of the receiving antenna in meters; $r$ the distance between the antennas in meters; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the urban density. \citeauthor{hata-propagation-loss} found this model to be representative for frequencies $f_c \sim$~150--1500~MHz, with transmitter heights $h_b \sim$~30--200~m, receiver heights $h_m \sim$~1--10~m and over distances $r\sim$~1--20~km.

Converting from a subtracted term in decibels to a flux density factor $L_S$ results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \frac{1}{10} 10^{L_p} = \zeta r^\eta,
\end{equation}
with $\eta$ and $\zeta$ given by
\begin{align} 
 \label{eq:zeta-definition} \eta & = 4.49 - 0.655 \log_{10} h_b,\\
 \label{eq:eta-definition}  \zeta & = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{align}
Note that according to Hata's model, the exponent of the power law $\eta$ depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. To find the rate density function $f_p$ that considers propagation effects, one can replace $S(r)$ in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{Ig}{4\pi \zeta r^\eta}.
\end{equation}
The resulting rate density function $f_p$ is
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(rho*%pi*r^2,S),solve(S(r)=S,r);
\begin{align} \label{eq:frequency-density-with-propagation}
f_p(S)
= & \frac{d}{dS} \left[ N - \rho \pi \left( \frac{Ig }{4\pi\zeta S} \right)^{2/\eta} \right]
= \frac{\rho 2\pi}{\eta S} \left( \frac{Ig}{4\pi\zeta S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free-space propagation effects, the observed log-log histogram is predicted to have a $-(\frac{2}{\eta}+1)$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram. According to Hata's model, this is valid for the range 30--200 m. The trend of the slope will not continue indefinitely when increasing the height further. Instead it will converge to a $-2$ slope, which corresponds to free-space propagation.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

We note that a uniform distribution of meteors or aircrafts which reflect free-space propagating RFI can create a power-law distribution with a similar slope: a uniform two-dimensional distribution of reflecting sources will create a $-1.5$ slope, while a uniform three-dimensional distribution will create a $-1.75$ slope. With brightness-distribution analyses one can therefore not distinguish between transmitters affected by Hata's propagation model and reflectors affected by free-space propagation. Reflected RFI might become relevant at lower amplitude levels.

@@

So far, these are the expected histograms for pure noise and pure RFI that propagates through free space. However, the measured distribution is a mixture of the two. Analytic derivation of the corresponding mixed amplitude distribution is not trivial, but the distributions can easily be estimated by drawing complex samples from the two distributions and calculating and counting the amplitudes. A sample can be drawn from the RFI distribution by integration, scaling and inversion of the rate density function in Eq.~\eqref{eq:frequency-density-with-propagation}. To invert the cumulative function, one needs to assume that there are no sources beyond some limiting distance $r_L$. With this assumption, a single complex RFI contaminated sample $S_{\textrm{RFI}}$ can be sampled with:
\begin{equation} \label{eq:drawing-rfi-with-propagation}
% === Proof of formula below ===
% From definition above:
%   F(S) = N-\rho \pi ( Ig / 4 \pi \zeta S ) ^ (2/\eta)
% We want S to be in the interval [ S_l ; \inf ], where: (from equation of S(r) above)
%   S_l = I g / 4 \pi \zeta r_l^\eta,
% the flux density of the most distance sources at distance r_l that are still visible.
% This means F(S) will be in the interval [ F(\inf) ; F(S_l) ], and:
%   F(\inf) = N
%    F(S_l) = N - \rho \pi r_l^2
% Because F is the CDF, we can sample a strength by calculating the S
% from a uniformly distributed F within this interval.
% If x_u is a uniform variable between 0 and 1, then N - x_u \rho \pi r_l^2 is a uniform
% variable in F's interval. From this we calculate S with the definition of F(S) :
%   N - x_u \rho \pi r_l^2 = N-\rho \pi ( Ig / 4 \pi \zeta S ) ^ (2/\eta)
%       x_u \rho \pi r_l^2 = \rho \pi ( Ig / 4 \pi \zeta S ) ^ (2/\eta)
%                x_u r_l^2 = ( Ig / 4 \pi \zeta S ) ^ (2/eta)
%   (x_u r_l^2 ) ^ (eta/2) = Ig / 4 \pi \zeta S
 %   x_u^(\eta/2) r_l^\eta = Ig / 4 \pi \zeta S
%                        S = Ig / ( 4 \pi \zeta x_u^(\eta/2) r_l^\eta )
S_{\textrm{RFI}} \leftarrow \frac{I g}{4\pi \zeta x_u^{\eta/2} r_L^\eta} e^{i 2 \pi y_u}.
\end{equation}
Here, $S_{\textrm{RFI}}$ is a new complex RFI sample that follows a power-law distribution; $\eta$ and $\zeta$ are defined in Eqs.~\eqref{eq:zeta-definition} and \eqref{eq:eta-definition}; $I$ is the average intrinsic strength; $g$ is the gain of the instrument; $0 < x_u,y_u \le 1$ are two independently drawn uniformly distributed samples; and $r_L$ is the maximum distance of visible sources. A sample $S$ that is contaminated by both RFI and noise can be drawn with
$S \leftarrow v_n + w_n i + S_{\textrm{RFI}}$, with $v_n, w_n \sim N(\mu=0; \sigma)$. An example of distribution curves of $S$ for $\eta=2$ and various settings of $Ig/4\pi \zeta r_L^2$ is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.

@@

From Eq.~\eqref{eq:frequency-density-with-propagation} it can be seen that the $\rho$, $I$ and $g$ parameters have the same effect of scaling the power-law distribution, and do not change its shape or slope. Therefore, with distribution analyses one can e.g. not determine whether the distribution is dominated by low-power sources within the horizon or by scattered signals from over the horizon. The horizon of an antenna is estimated with $\sqrt{2hr}$ \citep{bullington-propagation}, with $r$ the radius of the Earth and $h$ the height of the antenna. For LOFAR, the horizon is at about 5~km.

@@

Fitting straight lines to the distribution curve in a log-log plot is not the most accurate way of estimating the exponent of a power-law distribution \citep{power-law-distribution}. However, because of our enormous sample size, which allows fitting the line over a large interval, the estimator will be sufficiently accurate for our purpose. Nevertheless, we will additionally calculate a maximum-likelihood estimator for comparison. The maximum-likelihood estimator for the exponent in a power-law distribution is given by the Hill estimator $\hat \alpha_H$ \citep{hill-estimator, power-law-distribution}, defined as:
\begin{equation} \label{eq:hill}
 \hat \alpha_H = 1 + N \left(\sum\limits_{i=1}^{N} \ln \frac{x_i}{x_\textrm{min}} \right)^{-1},
\end{equation}
with $N$ the number of samples and $x_i$ for $0 < i \le N$ the samples that follow a power law with lower bound $x_\textrm{min}$.

\begin{figure}
\begin{center}
\includegraphics[width=7.5cm]{img/explanation-lower-constraint/FindingConstraints}
\caption{Cartoon of how a constraint on the lower fall-off point of the power-law distribution can be determined. Note that the labelled areas are areas as occupied in a linear plot, i.e., the integration of the density function. Areas in a log-log plot are not linearly related to the integral. There are two ways to estimate the lower constraint $S_L$: (i) the areas $N_a$ and $N_\textrm{total}-N_\textrm{part}$ are equal if $Ig/r_L^\eta$ is constant during the observation, and (ii) if one assumes $Ig/r_L^\eta\sim$ uniform, then $N_a+N_b = N_\textrm{total}-N_\textrm{part}$.}
\label{fig:explanation-constraints}
\end{center}
\end{figure}

\subsection{Determining RFI distribution limits} \label{sec:rfi-distribution-constraints}
In this section we will show how to put upper and lower constraints on the power-law distribution. Assume that we have found a power law with exponent $\alpha$ and factor $\beta$ over an amplitude region $[S_1 ; S_2]$, resulting in the rate density function $h(S) = \beta S^\alpha$. $S_1$ and $S_2$ are selected by visual inspection of the histogram. Assume that the histogram contains $N_\textrm{part}$ (RFI) samples with amplitude $> S_1$, as sketched in Fig.~\ref{fig:explanation-constraints}, and that the effect of the Rayleigh component on the histogram $> S_1$ is negligible. The hypothetical upper limit $S_U$ of the distribution can be found by solving
\begin{equation} \label{eq:upper-limit-constrained}
 \int\limits_{S_1}^{S_U} h(S) dS = N_\textrm{part}.
\end{equation}
The observed histogram will break down beyond some amplitude because of several reasons: the samples are digitized with an analogue-to-digital converter (ADC) with limited range; we observe for a limited time and the rate count is discrete; and, under the assumption of a uniform spatial distribution of RFI transmitters, samples with very high amplitude would have to be produced by transmitters that are very close to the telescope. However, it is likely that the uniform spatial distribution of transmitters will break down at closer distances.
